|

### 2025-11-07 manuscript/paper.Rmd
fucking citation.
paper update.

Consider a set of $n$ teams indexed from $i = 1, 2, \cdots, n$ each with
an associated true strength parameter
$\boldsymbol{\theta} = (\theta_1, \theta_2, \ldots, \theta_n)$ such that
$(\theta_1 > \theta_2 > \ldots > \theta_n)$. Next, let
$r(\boldsymbol{\theta})$ be the indexes of $\boldsymbol{\theta}$ when
$\theta$ is sorted from largest to smallest so that
$r(\boldsymbol{\theta}) = \{1, 2, \ldots, n-1, n\}$. Next define
$T(\phi, s)$ as a tournament with schedule $\phi$ and seeding structure
$s$. $\phi$ contains all the information about the scheduling of teams,
which could be fully known prior to the tournament (*e.g.* round robin)
or determined as the tournament progresses (*e.g.* single elimination
tournament). We then let $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ be an
$n$-dimensional vector-valued random variable that gives the results of
a tournament as a vector of the indexes of the vector
$\boldsymbol{\theta}$. The index in the first position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished first, the index in the second position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished second, and so on.

@@

If one views the true ranking of the teams in a tournament as a message
to be sent to a receiver and the outcome of the tournament as a message
that is received, we can measure the "goodness" of a tournament in terms
of its ability to accurately transmit the true ranking. We can then use
concepts from information theory to assess the ability of a tournament
to correctly rank teams. Specifically, we start with the concept of
mutual information (@Guiasu1977-nu). For two random variables $X$ and
$Y$ mutual information is defined as:

@@

We note here that $r$ is not a random variable, however, since the
indexing of the teams in the vector $\boldsymbol{\theta}$ is arbitrary
(we use 1, 2, 3, etc. for convenience, but any indexing is valid), we
can view this a random variable where all permutation so of the $n$
teams are equally likely. Therefore when calculating the mutual
information in this setting, we set $p(r) = \frac{1}{n!}$. By a similar
argument we set $p(\hat{r}) = \frac{1}{n!}$.

@@

$$
I(r,\hat{r})= n!\sum_{\hat{r}}p(r,\hat{r})\log \frac {p(r,\hat{r})}{\left(\frac{1}{n!}\right)^2} = n! log((n!)^2)
$$ and for the specific case when $n=4$ would be: $$
4! log_2(4!^2) = 24*log_2(24^2) = 220.0782
$$
# Weighted Mutual Information

In order to alleviate this problem, we instead consider *weighted* mutual information (@Guiasu1977-nu). We want to give more weight to permutations from $\hat{r}$ that are "closer" to $r$.  Essentially, this problem comes down to choosing a metric that measures the distance between the vectors $\hat{r}$ and $r$. One then gives more weight to combinations of $\hat{r}$ and $r$ that are closer and less weight to combinations that are far apart.  A natural choice for distance between vectors the sum of squared deviations between the two vectors.  Specifically, we define the loss function between $\hat{r}$ and $r$ to be $l(r, \hat{r}) = \sum_{i=1}^n (\hat{r}_i - r_i)^2$.  Then, since small losses should be associated with large weights, we focus on the recipricol of this loss function.

@@

As an example, consider $r = (1,2,3,4)$ and $\hat{r} = (1,3,4,2)$.  This requires two adjacent swaps to get $\hat{r}$ to equal $r$ (i.e swap 4 with 2 and then 3 with 2). The Kendall's $\tau$ value between these two vectors is `r cor(c(1,2,3,4),c(1,3,4,2), method = "kendall")`.  We would then define the weight between the two vectors to be $(\rho_{\tau} + 1)/2 =$ `r (cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") + 1)/2`.  This is exactly equivalent to $1 - \frac{\#\text{ adj swaps}}{{n \choose 2}}$.

@@

@@

@shieh1998weighted: Weighted Kendall's Tau

@@

$$
w(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) = \sum_{i = 1}^n m_i \times d(\hat{r}_i, r_i)

\\

wI_m(r_{\{{i|m_i >0}\}}, \hat{r_{\{{i|m_i >0}\}}})=\sum_{r_{\{{i|m_i >0}\}}}\sum_{\hat{r}_{\{{i|m_i >0}\}}} w(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) \times \\
p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})\times \\ \log \frac {p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})}{p(r_{\{{i  \mid m_i >0}\}})p(\hat{r}_{\{{i \mid m_i >0}\}})}
$$

Because this mutual information is unitless, the formula can be
standardized to be between 0 and 1 as follows: $$
\frac{wI_m(r,\hat{r})}{\text{max\{}H(r), H(\hat{r})\}}
$$

>>>>

Consider a set of $n$ teams indexed from $i = 1, 2, \cdots, n$ each with
an associated true strength parameter
$\boldsymbol{\theta} = (\theta_1, \theta_2, \ldots, \theta_n)$ such that
$(\theta_1 \ge \theta_2 \ge \ldots \ge \theta_n)$. Next, let
$r(\boldsymbol{\theta})$ be the indexes of $\boldsymbol{\theta}$ when
$\theta$ is sorted from largest to smallest so that
$r(\boldsymbol{\theta}) = \{1, 2, \ldots, n-1, n\}$ (Note: While we allow for two values of $\theta$ to be equal to each other, we choose to assign unique ranks to each $\theta_i$).  Next define
$T(\phi, s)$ as a tournament with schedule $\phi$ and seeding structure
$s$. $\phi$ contains all the information about the scheduling of teams,
which could be fully known prior to the tournament (*e.g.* round robin)
or determined as the tournament progresses (*e.g.* single elimination
tournament). We then let $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ be an
$n$-dimensional vector-valued random variable that gives the results of
a tournament as a vector of the indexes of
$\boldsymbol{\theta}$. The number in the first position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished first, the number in the second position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished second, and so on.  (Note: We do allow for ties in $\hat{r}$.)

@@

As noted above we allow for ties in $\hat{r}$.  Continuing the above example, if $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ = (3, 2, 1, 3) this indicates that the teams that are truly ranked 1st and 4th tied for 3rd.

We need a transition: 
If one views the true ranking of the teams in a tournament as a message
to be sent to a receiver and the outcome of the tournament as a message
that is received, we can measure the "goodness" of a tournament in terms
of its ability to accurately transmit the true ranking. We can then use
concepts from information theory to assess the ability of a tournament
to correctly rank teams. Specifically, we start with the concept of
mutual information (@Guiasu1977-nu).

For two random variables $X$ and $Y$ mutual information is defined as:

@@

In order to compute mutual information, marginal distributions of $r$ and $\hat{r}$ are required.  Without knowing $r$, all possible permutations of $\hat{r}$ are equally likely so we set $p(\hat{r}) = \frac{1}{n!}$. By a similar argument we set $p(r) = \frac{1}{n!}$.

we set . By a similar argument we set $p(\hat{r}) = \frac{1}{n!}$.

@@

$$
I(r,\hat{r})= n!\sum_{\hat{r}}p(r,\hat{r})\log \frac {p(r,\hat{r})}{\left(\frac{1}{n!}\right)^2} = n! log((n!)^2)
$$ and for the specific case when $n=4$ would be: 
$$
4! log_2(4!^2) = 24*log_2(24^2) = 220.0782
$$

# Weighted Mutual Information

In order to alleviate this problem, we instead consider *weighted* mutual information (@Guiasu1977-nu). We want to give more weight to permutations from $\hat{r}$ that are "closer" to $r$.  Essentially, this problem comes down to choosing a metric that measures the closeness between the vectors $\hat{r}$ and $r$. One then gives more weight to combinations of $\hat{r}$ and $r$ that are closer and less weight to combinations that are far apart.  A natural choice for distance between vectors the sum of squared deviations between the two vectors.  Specifically, we define the loss function between $\hat{r}$ and $r$ to be $l(r, \hat{r}) = \sum_{i=1}^n (\hat{r}_i - r_i)^2$.  Then, since small losses should be associated with large weights, we focus on the recipricol of this loss function.

@@

As an example, consider $r = (1,2,3,4)$ and $\hat{r} = (1,3,4,2)$.  This requires two adjacent swaps to get $\hat{r}$ to equal $r$ (i.e swap 4 with 2 and then 3 with 2). The Kendall's $\tau$ value between these two vectors is $\frac{1}{3}$.  We would then define the weight between the two vectors to be $(\rho_{\tau} + 1)/2 = \frac{2}{3}$ .  This is exactly equivalent to $1 - \frac{\#\text{ adj swaps}}{{n \choose 2}}$.

<!-- `r cor(c(1,2,3,4),c(1,3,4,2), method = "kendall")` -->
<!-- `r (cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") + 1)/2` -->

@@

One and appropriate weight function was chosen, we get the following formula for weighted information:

$$
I_w(r,\hat{r})= \sum_{r}\sum_{\hat{r}}w(r, \hat{r})p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$

@@

@Shieh1998: Weighted Kendall's Tau

@@

$$
wI_m(r_{\{{i|m_i >0}\}}, \hat{r}_{\{{i|m_i >0}\}})=\sum_{r_{\{{i|m_i >0}\}}}\sum_{\hat{r}_{\{{i|m_i >0}\}}} w_m(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) \times
$$

<!-- $$ -->
<!-- p(\hat{r}_{\{{i|m_i > 0}\}}, r_{\{{i|m_i >0}\}})\times \\  -->

<!-- \log \frac{p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})}{p(r_{\{{i  \mid m_i >0}\}})p(\hat{r}_{\{{i \mid m_i >0}\}})} -->
<!-- $$ -->

Because this mutual information is unitless, the formula can be standardized to be between 0 and 1 as follows: 
$$
\frac{wI_m(r,\hat{r})}{\text{max}\{H(r), H(\hat{r})\}}
$$
